Progress Report on One Simulated Dataset

### 2023-11-09 Docs/ProgressReportSim.Rmd
pred difference from scaled or unscaled parameters

---
title: "Progress Report on One Simulated Dataset"
author: "Ying Jin"
date: "2023-09-15"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

@@

```{r}
# generated testing data 
N_test <- 200
b_test <- matrix(rnorm(N_test*K), N_test, K) %*% diag(sqrt(lambda)) %*% t(phi) # of size N by J
# dim(b_test)

# latent gaussian function
eta_test <- t(vapply(1:N_test, function(x){
    f_0(x) + b_i[x,]
  }, numeric(J)))

@@

# format into dataframe
test_df <- data.frame(id = factor(rep(1:N_test, each=J)),
                 t = rep(t, N_test), 
                 Y = as.vector(t(Y_test)),
                 eta_i = as.vector(t(eta_test)),
                 sind = rep(1:J, N_test))
# bin
test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
test_df$bin <- as.numeric(as.character(test_df$bin))
test_df$id <- as.factor(test_df$id)
test_id <- unique(test_df$id)
```

@@

```{r}
# the model
## now, with expanded mean and eigenfuncitons, a Bernoulli distribution at each time point
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

@@

```{r}
# parameter names
name_lst <- as.list(rep(0, K))
names(name_lst) <- paste("xi", 1:K, sep = "")
parm.names <- as.parm.names(name_lst)
pos.xi <- grep("xi", parm.names)
PGF <- function(Data) {
  xi <- rnorm(Data$K)
  return(xi)
}
mon.names <- "LP"
```

```{r pred_debias_unscale, results='hide'}
# approximation process
# container
pred_list <- list()
# score <- array(NA, dim=c(length(test_id), K, 4))
# dimensions: subject, eigenfunction, maximum observation time

# prediction for a single subject
t1 <- Sys.time()
# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_new[1:max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

@@

<!-- ### Scaled eigenfunctions and eigenvalues -->

```{r pred_debias_scale, results='hide'}
## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
# container
pred_list3 <- list()

# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

@@

    # into a list
    MyData <- list(K=K, PGF=PGF, 
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

@@

  pred_list3[[i]] <- df_i
}
```

@@

- The scaled and unscaled parameters seem to make a difference in prediction, which they shouldn't

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2, 2))
plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5, 
     xlab="unscaled", ylab="scaled", main = "Up to 0.2")
plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.4")
plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.6")
plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
     xlab="unscaled", ylab="scaled", main = "Up to 0.8")
```

```{r}
data.frame(
  df_pred$pred0.4[df_pred$t>0.4],
  df_pred3$pred0.4[df_pred3$t>0.4]
)

@@

```{r, class.source='fold-show'}
# model fit
t1 <- Sys.time()
fit_adglmm <- mixed_model(Y ~ t, random = ~ t | id, data =  train_df, family = binomial())
t2 <- Sys.time()
# time 
t2-t1
# model
summary(fit_adglmm)
```

@@

```{r, class.source = "fold-show"}
# subset 10 subjects from training dataset 
df_train_sub <- train_df %>% filter(id %in% 1:10 & sind==bin)
df_train_sub$id <- droplevels(df_train_sub$id)
# levels(df_train_sub$id)
# intersect(t, unique(df_train_sub$t))
```

@@

```{r, class.source="fold-show"}
# construct design matrix for the fixed effect intercept
t_sub <- unique(df_train_sub$t)
fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
dim(fix_xmat[[1]]$X)

# basis functions
fix_xmat <- data.frame(t=t_sub, 
                       fix_xmat[[1]]$X)
colnames(fix_xmat) <- c("t", paste0("B", 1:5))

@@

```{r, class.source = "fold-show"}
# construct design matrix for the random effect
rand_xmat <- smoothCon(s(t, bs="cr", k=4), data = data.frame(t=t_sub))

dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
rand_xmat <- data.frame(t = t, rand_xmat[[1]]$X)

@@

The model fitting took about 6 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.

@@

```{r, eval=TRUE, class.source = "fold-show"}
# model fit
t1 <- Sys.time()
fit_adglmm <- mixed_model(fixed = Y ~ 0+B1+B2+B3+B4+B5, 
                          random = ~X1+X2+X3+X4 | id,
                          data =  df_train_sub, family = binomial())
t2 <- Sys.time()
t2-t1

# model
summary(fit_adglmm)
```

@@

```{r}
df_test_sub <- test_df %>% filter(id %in% 1:10 & sind==bin)
df_test_sub$id <- droplevels(df_test_sub$id)
# levels(df_test_sub$id)

@@

colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:15], "pred1", "pred2", "pred3", "pred4")
```

@@

>>>>

---
title: "Progress Report on One Simulated Dataset"
author: "Ying Jin"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

@@

```{r}
# generated testing data 
N_test <- 200
# xi_test <- matrix(rnorm(N_test*K), N_test, K)
xi_test <- rmvnorm(N_test, mean = rep(0, K), sigma = diag(lambda))
b_test <- xi_test %*% t(phi) # of size N by J
# dim(b_test)

# latent gaussian function
eta_test <- t(vapply(1:N_test, function(x){
    f_0(x) + b_test[x,]
  }, numeric(J)))

@@

# format into dataframe
test_df <- data.frame(id = factor(rep(1:N_test, each=J)),
                 t = rep(t, N_test), 
                 Y = as.vector(t(Y_test)),
                 eta_i = as.vector(t(eta_test)),
                 sind = rep(1:J, N_test))
# bin
# test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
# test_df$bin <- as.numeric(as.character(test_df$bin))
test_df$id <- as.factor(test_df$id)
test_id <- unique(test_df$id)
```

@@

```{r model}
# the model
## now, with expanded mean and eigenfuncitons, a Bernoulli distribution at each time point
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

@@

```{r}
# parameter names
name_lst <- as.list(rep(0, K))
names(name_lst) <- paste("xi", 1:K, sep = "")
parm.names <- as.parm.names(name_lst)
pos.xi <- grep("xi", parm.names)
# PGF <- function(Data) {
#   xi <- rnorm(Data$K)
#   return(xi)
# }
mon.names <- "LP"
```

```{r pred_debias_unscale, results='hide'}
# approximation process
# container
pred_list <- list()
converge_vec <- matrix(NA, N_test, 4)
# score <- array(NA, dim=c(length(test_id), K, 4))
# dimensions: subject, eigenfunction, maximum observation time

# prediction for a single subject
t1 <- Sys.time()
# per subject
# seq_along(test_id)
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

@@

    # into a list
    MyData <- list(K=K, 
                   X=efunctions_new[1:max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS", Iterations = 1000,
                                CovEst = "Identity")
    converge_vec[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

@@

```{r pred_debias_scale, results='hide'}
## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
# container
pred_list3 <- list()
converge_vec2 <- matrix(NA, N_test, 4)
# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

@@

    # into a list
    MyData <- list(K=K,
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS", Iterations = 1000,
                                CovEst = "Identity")
    converge_vec2[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
    score <- Fit$Summary1[, "Mode"]

@@

  pred_list3[[i]] <- df_i
}

mean(converge_vec2)
```

@@

When using the unscaled estimates of eigenvalues and eigenfunctions, the Laplace Approximation does not converge all the time (convergence rate = `r round(mean(converge_vec), 2)`). And when using scaled estimates, the convergence rate is even lower (`r round(mean(converge_vec2), 2)`).

When I increasing the maximum iteration to 1000, the predictive performance from two sets of estiamtes got very similar. However, the convergence rate is still very different. Below I try to show a example of the discrepancy in prediction. As we see, the most problematic situation is when observations is up to 0.4.

```{r}
# subjects with higher deiviation 
# subjects with greatest difference
id_diff <- df_pred %>% select(id, t, pred0.4) %>%
  full_join(df_pred3 %>% select(id, t, pred0.4), by = c("id", "t")) %>% 
  group_by(id) %>%
  summarise(diff = mean((pred0.4.x-pred0.4.y)^2, na.rm = T)) %>% 
  arrange(desc(diff))

# subject 43, 79, 57, 148 have the most different prediction
df_pred %>% filter(id == id_diff$id[1]) %>% 
  select(t, starts_with("pred")) %>%
  pivot_longer(starts_with("pred")) %>%
  left_join(df_pred3 %>% filter(id == id_diff$id[1]) %>%
              select(t, starts_with("pred")) %>%
              pivot_longer(starts_with("pred")),
            by = c("t", "name")) %>%
  ggplot()+
  geom_point(aes(x=value.x, y=value.y), na.rm = T)+
  facet_wrap(~name)+
  labs(title = paste0("Subject ", id_diff$id[1]), x="Unscaled", y="Scaled")
```

Think about the estimates. The scaled estimates of eigenvalues are much smaller than the unscaled once, meaning the prior distribution of score using scaled estimates would be much more centered thant the unscaled ones. This would also drive the posterior distribution to be in similar shapes. Since the distribution from scaled eigenvalues has such high density at certain values, could it be that the sampling quickly got trapped in a small interval that is far from the true optimal values?

Take the same subject with observation up to 0.4 as an example. Below is the iterative history until convergence. In this case, unscaled estiamtes converged at the 67 iteration and the scaled one did not converge.

```{r, results='hide'}
this_id <- id_diff$id[1]
this_t <- 0.4

# data
df_it <- test_df %>% filter(id==this_id & t <= this_t)
max_rid <- nrow(df_it)

#### use Unscaled params ####
MyData_unscaled <- list(K=K, 
               X=efunctions_new[1:max_rid, ], 
               mon.names=mon.names,
               parm.names=parm.names, 
               pos.xi=pos.xi, 
               y=df_it$Y, 
               tao=diag(new_lambda), f0=new_mu[1:max_rid])
Fit_unscaled <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData,
                                     Method = "BFGS", Iterations = 1000,
                            CovEst = "Identity")

#### Scaled params ####
MyData_scaled <- list(K=K, 
               X=efunctions_scaled[1:max_rid, ], 
               mon.names=mon.names,
               parm.names=parm.names, 
               pos.xi=pos.xi, 
               y=df_it$Y, 
               tao=diag(scaled_lambda), f0=new_mu[1:max_rid])
Fit_scaled <- LaplaceApproximation(Model, parm = rep(0, K), 
                                   Data=MyData_scaled, Method = "BFGS", 
                                   Iterations = 1000,
                            CovEst = "Identity")

## iterative distribution, compared to true value
true_xi_it <- data.frame(name = colnames(Fit_unscaled$History),
                         true_xi = xi_test[id_diff$id[1], ])
```

```{r}
ggarrange(
Fit_unscaled$History %>% as.data.frame() %>%
  mutate(iter = 1:nrow(Fit$History)) %>%
  pivot_longer(starts_with("xi")) %>%
  left_join(true_xi_it, by = "name") %>%
  mutate(value = value/sqrt(n_bin)) %>%
  mutate_at(vars(value, true_xi), abs) %>%
  ggplot()+
  geom_point(aes(x=iter, y=value))+
  geom_line(aes(x=iter, y = true_xi, col = "red"))+
  facet_wrap(~name)+
  labs(title = "Unscaled parameter"),

Fit_scaled$History %>% as.data.frame() %>%
  mutate(iter = 1:nrow(Fit_scaled$History)) %>%
  pivot_longer(starts_with("xi")) %>%
  left_join(true_xi_it, by = "name") %>%
  mutate_at(vars(value, true_xi), abs) %>%
  ggplot()+
  geom_point(aes(x=iter, y=value))+
  geom_line(aes(x=iter, y = true_xi, col = "red"))+
  facet_wrap(~name)+
  labs(title = "Scaled parameter"), nrow = 1, common.legend = T)
```

Let's zoom in on the later part of the scaled parameter. It looks like sampling bounced symmetrically around a certain line.

```{r}
Fit_scaled$History %>% as.data.frame() %>%
  mutate(iter = 1:nrow(Fit_scaled$History)) %>%
  filter(iter >= 100) %>%
  pivot_longer(starts_with("xi")) %>%
  left_join(true_xi_it, by = "name") %>%
  mutate_at(vars(value, true_xi), abs) %>%
  ggplot()+
  geom_point(aes(x=iter, y=value), size = 0.2)+
  # geom_line(aes(x=iter, y = true_xi, col = "red"))+
  facet_wrap(~name, scales = "free")+
  labs(title = "Scaled parameter, after 100th iteration")
```

@@

```{r}
# model fit
t1 <- Sys.time()
fit_adglmm <- mixed_model(Y ~ t, random = ~ t | id, data =  train_df, family = binomial())
t2 <- Sys.time()
# t2-t1
```

@@

```{r}
# subset 10 subjects from training dataset 
df_train_sub <- train_df %>% filter(id %in% 1:10 & sind==bin)
df_train_sub$id <- droplevels(df_train_sub$id)
# levels(df_train_sub$id)
# intersect(t, unique(df_train_sub$t))
# View(df_train_sub)
```

@@

```{r}
# construct design matrix for the fixed effect intercept
t_sub <- unique(df_train_sub$t)
# fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
fix_xmat <- ns(t_sub, df=5)
# dim(fix_xmat[[1]]$X)

# basis functions
fix_xmat <- data.frame(t=t_sub, 
                       fix_xmat)
colnames(fix_xmat) <- c("t", paste0("B", 1:5))

@@

```{r}
# construct design matrix for the random effect
rand_xmat <- smoothCon(s(t, bs="cr", k=4), data = data.frame(t=t_sub))
# rand_xmat <- ns(t_sub, df = 4)

# dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
rand_xmat <- data.frame(t = t, rand_xmat[[1]]$X)

@@

The model fitting took about 12 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.

@@

```{r, eval=TRUE, class.source = "fold-show"}
# model fit
t1 <- Sys.time()
fit_adglmm <- mixed_model(fixed = Y ~ 0+B1+B2+B3+B4+B5, 
                          random = ~X1+X2+X3+X4 | id,
                          data =  df_train_sub, family = binomial())
t2 <- Sys.time()
# t2-t1

@@

```{r}
df_test_sub <- test_df %>% filter(id %in% 1:10 & sind %in% mid)
df_test_sub$id <- droplevels(df_test_sub$id)
# levels(df_test_sub$id)

@@

colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:14], "pred1", "pred2", "pred3", "pred4")
```

@@

# Notes

Perhaps I should try to verify my thoughts about laplace approximation by different generation scheme. For example, what if the true eigenvalues are on larger scale?
